---
title: "Movielens"
author: "Lakshmi Veeramani"
date: "5/1/2021"
output:
  pdf_document: Movielens.pdf
  RMD_document: Movielens.Rmd
---



---
## Project overview Movie recommendation system

Dataset used: "http://files.grouplens.org/datasets/movielens/ml-10m.zip"
Data Description: The dataset is named as movielens dataset, This contains
100 million movies , rating, users and Genres information.

Scope of the Project: Develop movie recommendation system for the movies that are not rated, using the information available like,User's rating, movie rating and genres rating.

Output: Rating to be predicted by the model developed by the project.

Metrics requirements: RMSE < 0.86490

I have tried movie rating average and user rating average and genres based models.
Then applied regularization on individual models and then combined models.
Finally tried the matrix factorization technique, which provided the lowest RMSE for the problem.
---


## Load the first setup code given in the course

```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}

##########################################################
# Create edx set, validation set (final hold-out test set)
##########################################################

## Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)

## MovieLens 10M dataset:
## https://grouplens.org/datasets/movielens/10m/
## http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 3.6 or earlier:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                            title = as.character(title),
                                            genres = as.character(genres))
# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                            title = as.character(title),
                                            genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
      semi_join(edx, by = "movieId") %>%
      semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```
## ###########End of code provided in the course #######################

## Exploratory Data Analysis#########################
```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
## Import libraries
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(dplyr)) install.packages("dplyr", repos = "http://cran.us.r-project.org")
if(!require(kableExtra)) install.packages("kableExtra", repos = "http://cran.us.r-project.org")
if(!require(tidyr)) install.packages("tidyr", repos = "http://cran.us.r-project.org")
if(!require(stringr)) install.packages("stringr", repos = "http://cran.us.r-project.org")
if(!require(forcats)) install.packages("forcats", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")
library(lubridate)
library(dplyr)
library(kableExtra)
library(tidyr)
library(stringr)
library(forcats)
library(ggplot2)

## Exploratory Data Analysis
glimpse(edx)
str(edx)
summary(edx)
names(edx)
head(edx, n=10)
nrow(edx)

#Check validation set 
head(validation)
nrow(validation)

#How many distinct movies, users and genres
message("Total Unique Movies: ", n_distinct(edx$movieId))
message("Total Unique Genres: ", n_distinct(edx$genres))
message("Total Unique Users: ", n_distinct(edx$userId))

#How many ratings?

message("Total Unique Ratings: ", n_distinct(edx$rating))
message("List of Unique Ratings: ", distinct(edx, rating))
```
## ###########################################################

```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
##Data Cleaning checks:  Are there any missing values? => No missing values
if (anyNA(edx) == 0){ 
  message("No missing values")}

## Data Cleaning steps:1.Timestamp to be converted to date time format

edx$ts_date <- as.POSIXct(edx$timestamp, origin="1970-01-01")
validation$ts_date <- as.POSIXct(validation$timestamp, origin="1970-01-01")

# Extract the rated year and rated month in Training data

edx$rated_year <- format(edx$ts_date,"%Y")
edx$rated_month <- format(edx$ts_date,"%m")

# Extract the rated year and rated month in validation data

validation$rated_year <- format(validation$ts_date,"%Y")
validation$rated_month <- format(validation$ts_date,"%m")

## Data Cleaning steps:2.Movie name has the movie release year, which needs to be splitted to have this info as separate column

edx <- edx %>% mutate(released_year = as.numeric(str_sub(title,-5,-2)))
validation <- validation %>% mutate(released_year = as.numeric(str_sub(title,-5,-2)))
head(edx, 2)

head(validation, 2)

# remove unnecessary columns on edx and validation dataset

edx <- edx %>% select(userId, movieId, rating, title, genres, released_year, rated_year, rated_month)

validation <- validation %>% select(userId, movieId, rating, title, genres, released_year, rated_year, rated_month)

head(edx, 2)

head(validation, 2)
```

```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
## ###########################################################
##Convert the columns in to desired datatypes
edx$rated_year <- as.numeric(edx$rated_year)
edx$rated_month <- as.numeric(edx$rated_month)
edx$released_year <- as.numeric(edx$released_year)
validation$rated_year <- as.numeric(validation$rated_year)
validation$rated_month <- as.numeric(validation$rated_month)
validation$released_year <- as.numeric(validation$released_year)
```

## Baseline Model

```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
#calculate the average of all movies
mu_hat  <- mean(edx$rating)

#Predict the RMSE on the validation set
RMSE_mean_model_result = RMSE(validation$rating, mu_hat)

results <- data.frame(model = "Project Goal", RMSE = 0.8649)

#Adding results to the results data frame that contains all RMSE results
results <- results %>% add_row(model = "Naive Mean baseline model", RMSE =RMSE_mean_model_result  )

print(results)
```

## Movie rating prediction based on the movie rating average

```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
#Calculate the average buy movie
movie_avgs <- edx %>%
              group_by(movieId) %>%
              summarize(b_i = mean(rating - mu_hat))

#compute the predicted ratings on the validation set
RMSE_movie_model <- validation %>%
                    left_join(movie_avgs, by = 'movieId') %>%
                    mutate(pred = mu_hat +b_i) %>%
                    pull(pred)

#Predict the RMSE on the validation set
RMSE_movie_model_result = RMSE(validation$rating, RMSE_movie_model)

#Adding results to the results data frame that contains all RMSE results
results <- results %>% add_row(model = "Movie based model", RMSE =RMSE_movie_model_result )
print(results)
```
## Movie rating prediction based on the movie+user rating average

```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
#Calculate the average by user

user_avgs <- edx %>%
             left_join(movie_avgs, by = 'movieId') %>%
             group_by(userId) %>%
             summarize(b_u = mean(rating - mu_hat - b_i))

#compute the predicted ratings on the validation set
RMSE_movie_user_model <- validation %>%
                    left_join(movie_avgs, by = 'movieId') %>%
                    left_join(user_avgs, by = 'userId') %>%
                    mutate(pred = mu_hat +b_i+b_u) %>%
                    pull(pred)

#Predict the RMSE on the validation set
RMSE_movie_user_model_result = RMSE(validation$rating, RMSE_movie_user_model)

#Adding results to the results data frame that contains all RMSE results
results <- results %>% add_row(model = "Movie user based model", RMSE =RMSE_movie_user_model_result )
print(results)
```
## Movie rating prediction based on the movie+user+genres rating average
```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
genre_pop <- edx %>%
             left_join(movie_avgs, by = 'movieId') %>%
             left_join(user_avgs, by = 'userId') %>%
             group_by(genres) %>%
             summarize(b_u_g = mean(rating - mu_hat - b_i-b_u))
             

#compute the predicted ratings on the validation set
RMSE_movie_user_genres_model <- validation %>%
                    left_join(movie_avgs, by = 'movieId') %>%
                    left_join(user_avgs, by = 'userId') %>%
                    left_join(genre_pop, by = 'genres') %>%
                    mutate(pred = mu_hat +b_i+b_u+b_u_g) %>%
                    pull(pred)

#Predict the RMSE on the validation set
RMSE_movie_user_genres_model_result = RMSE(validation$rating, RMSE_movie_user_genres_model)

#Adding results to the results data frame that contains all RMSE results
results <- results %>% add_row(model = "Movie user genre based model", RMSE =RMSE_movie_user_genres_model_result )
print(results)
```
## Regularised movie based model
```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
lambdas <- seq(0, 15, 1)

rmses <- sapply(lambdas, function(lambda) {
    
    mu_hat  <- mean(edx$rating)
    # Calculate the avg by movieId
    b_i <- edx %>%
           group_by(movieId) %>%
           summarize(b_i = sum(rating - mu_hat)/(n()+lambda))


    #Compute the predicted ratings on validation set
    predicted_ratings <- validation %>%
                         left_join(b_i, by = 'movieId') %>%                                
                         mutate(pred = mu_hat +b_i) %>%
                         pull(pred)
    # Predict the RMSE on the validation set
    return(RMSE(validation$rating, predicted_ratings))
})

#Get the lambda value for the lowest RMSE
min_lambda <- lambdas[which.min(rmses)]

#Predict the RMSE on the validation set
regularised_movie_model <- min(rmses)

#Predict the RMSE on the validation set
RMSE_regularised_movie_model = min(rmses)

#Adding results to the results data frame that contains all RMSE results
results <- results %>% add_row(model = "Regularised Movie based model", RMSE =RMSE_regularised_movie_model )

print(results)

```
## Regularised user based model
```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
lambdas <- seq(0, 15, 1)

rmses <- sapply(lambdas, function(lambda) {
    
    mu_hat  <- mean(edx$rating)
    # Calculate the avg by movieId
    b_u <- edx %>%
           group_by(userId) %>%
           summarize(b_u = sum(rating - mu_hat)/(n()+lambda))


    #Compute the predicted ratings on validation set
    predicted_ratings <- validation %>%
                         left_join(b_u, by = 'userId') %>%                                
                         mutate(pred = mu_hat +b_u) %>%
                         pull(pred)
    # Predict the RMSE on the validation set
    return(RMSE(validation$rating, predicted_ratings))
})

#Get the lambda value for the lowest RMSE
min_lambda <- lambdas[which.min(rmses)]

#Predict the RMSE on the validation set
regularised_user_model <- min(rmses)

#Predict the RMSE on the validation set
RMSE_regularised_user_model = min(rmses)

#Adding results to the results data frame that contains all RMSE results
results <- results %>% add_row(model = "Regularised user based model", RMSE =RMSE_regularised_user_model )

print(results)
```
## Regularised movie + user based model
```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
lambdas <- seq(0, 15, 1)

rmses <- sapply(lambdas, function(lambda) {
    
    mu_hat  <- mean(edx$rating)
    # Calculate the avg by movieId
    b_i <- edx %>%
           group_by(movieId) %>%
           summarize(b_i = sum(rating - mu_hat)/(n()+lambda))
    # Calculate the avg by userId
    b_u <- edx %>%
           left_join(b_i, by='movieId')%>%
           group_by(userId) %>%
           summarize(b_u = sum(rating - b_i - mu_hat) / (n() + lambda))

    #Compute the predicted ratings on validation set
    predicted_ratings <- validation %>%
                         left_join(b_i, by = 'movieId') %>%
                         left_join(b_u, by = 'userId') %>%                         
                         mutate(pred = mu_hat +b_i+b_u) %>%
                         pull(pred)
    # Predict the RMSE on the validation set
    return(RMSE(validation$rating, predicted_ratings))
})

#Get the lambda value for the lowest RMSE
min_lambda <- lambdas[which.min(rmses)]

#Predict the RMSE on the validation set
regularised_movie_user_model <- min(rmses)

#Predict the RMSE on the validation set
RMSE_regularised_movie_user_model = min(rmses)

#Adding results to the results data frame that contains all RMSE results
results <- results %>% add_row(model = "Regularised Movie user based model", RMSE =RMSE_regularised_movie_user_model )
print(results)
```
## Regularised movie+user+genres based model
```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
lambdas <- seq(0, 15, 1)

rmses <- sapply(lambdas, function(lambda) {
    # Calculate the avg by movieId
    b_i <- edx %>%
           group_by(movieId) %>%
           summarize(b_i = sum(rating - mu_hat)/(n()+lambda))
    # Calculate the avg by userId
    b_u <- edx %>%
           left_join(b_i, by='movieId')%>%
           group_by(userId) %>%
           summarize(b_u = sum(rating - b_i - mu_hat) / (n() + lambda))
    # Calculate the avg by genre
    b_u_g <- edx %>%
           left_join(b_i, by='movieId')%>%
           left_join(b_u, by='userId')%>%
           group_by(genres) %>%
           summarize(b_u_g = sum(rating - b_i - mu_hat-b_u) / (n() + lambda))
    #Compute the predicted ratings on validation set
    predicted_ratings <- validation %>%
                         left_join(b_i, by = 'movieId') %>%
                         left_join(b_u, by = 'userId') %>%
                         left_join(b_u_g, by = 'genres') %>%
                         mutate(pred = mu_hat +b_i+b_u+b_u_g) %>%
                         pull(pred)
    # Predict the RMSE on the validation set
    return(RMSE(validation$rating, predicted_ratings))
})

#Get the lambda value for the lowest RMSE
min_lambda <- lambdas[which.min(rmses)]

#Predict the RMSE on the validation set
regularised_movie_user_genres_model <- min(rmses)

#Predict the RMSE on the validation set
RMSE_regularised_movie_user_genres_model = min(rmses)

#Adding results to the results data frame that contains all RMSE results
results <- results %>% add_row(model = "Regularised Movie user genres based model", RMSE =RMSE_regularised_movie_user_genres_model )
print(results)
```
Recommender system based on Reco package - Matrix factorization method

```{r, echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE}
if(!require(recosystem)) 
  install.packages("recosystem", repos = "http://cran.us.r-project.org")
set.seed(123, sample.kind = "Rounding") 

# Convert the edx and validation sets into recosystem input format

edx_reco <-  with(edx, data_memory(user_index = userId, 
                                           item_index = movieId, 
                                           rating     = rating))
validation_reco  <-  with(validation,  data_memory(user_index = userId, 
                                           item_index = movieId, 
                                           rating     = rating))

# Create the model 
r <-  recosystem::Reco()

# Choose the best tuning parameters
opts <- r$tune(edx_reco, opts = list(dim = c(10, 20, 30), 
                                       lrate = c(0.1, 0.2),
                                       costp_l2 = c(0.01, 0.1), 
                                       costq_l2 = c(0.01, 0.1),
                                       nthread  = 4, niter = 10))

 # Train the algorithm with edx_reco 
r$train(edx_reco, opts = c(opts$min, nthread = 4, niter = 20))

# Predict ratings  
y_hat_reco <-  r$predict(validation_reco, out_memory())

#Predict the RMSE on the validation set
RMSE_reco_matrix_factorization_model = RMSE(validation$rating, y_hat_reco)

#Adding results to the results data frame that contains all RMSE results
results <- results %>% add_row(model = "Reco matrix factorization model", RMSE =RMSE_reco_matrix_factorization_model )

results

```

